### 2023-10-29 tex/agda-primer.tex
[ writing ] ready for submission

I use the proof assistant and programming language Agda throughout this thesis,
with Agda code being used particularly in this chapter and
\cref{sec:framework,sec:semantics,sec:example-semantics}.
As such, it is important for the reader to be able to read basic Agda syntax in
order to benefit from the parts of the exposition that reside in code listings.
The syntax of Agda is broadly similar to that of Haskell, and relatively close
to that of Standard ML, OCaml, and Coq's Gallina sublanguage.
I will assume that the reader is able to read basic Haskell code, and spend most
time explaining differences thereof.

\subsection{Lexing and parsing}

@@

On its own, an underscore has a completely different meaning, which can depend
on context.
In patterns, an underscore has the same meaning as it has in Haskell and ML ---
it holds the place of a pattern variable, but does not name that variable.
In expressions, an underscore stands for an unspecified subterm which will be
solved by unification.
The solving of unspecified terms is canonical and respects $\beta\eta$-equality,
unlike in Coq.

@@

An unfortunate feature of the definition \AgdaFunction{id$_0$} is that we cannot
apply it to the expression \AgdaPrimitive{Set}, because \AgdaPrimitive{Set}
contains only small types, and itself is a large type.
We can work around these size issues using \emph{universe level polymorphism},
as in the following definition.

@@

Slightly more flexible than parameters are \emph{Protestant indices}%
\footnote{The terminology of Protestant/Catholic indices comes orally from Conor
McBride. I am not aware of a written reference.}.
Protestant indices also appear to the left of the colon, and also must appear
unmodified in the \emph{return type} of all of the constructors.
However, they may take different values in inductive appearances of the type
family in the argument types of constructors.
Protestant indices give a generalisation of polymorphic
recursion to indices of arbitrary type~\citep{Mycroft84,Henglein93}.

@@

The second example of a type family with Catholic indices is more general in
nature.
Below I define \emph{propositional equality}, written
\AgdaDatatype{\_$\equiv$\_}.
It has two parameters and one Catholic index (though the standard library
version of propositional equality I use throughout this thesis has an extra
level parameter for the sake of universe level polymorphism).
The constructor \AgdaInductiveConstructor{refl} constructs an inhabitant of
\AgdaArgument{M}\AgdaSpace{}\AgdaDatatype{\_$\equiv$\_}\AgdaSpace{}%
\AgdaArgument{N} only when \AgdaArgument{N} is definitionally equal to
\AgdaArgument{M} (because terms are considered ``the same'' to the type checker
exactly when they are definitionally equal).
Notice that \AgdaInductiveConstructor{refl} does not quantify over \AgdaBound{x}
because \AgdaBound{x} is already in scope as a parameter.

>>>>

I use the proof assistant and programming language Agda throughout this thesis,
with Agda code being used particularly in this chapter and
\cref{sec:framework,sec:semantics,sec:example-semantics}.
As such, it is important for the reader to be able to read basic Agda syntax in
order to benefit from the parts of the exposition that reside in code listings.
The syntax of Agda is broadly similar to that of Haskell~\citep{Haskell}, and
relatively close to that of Standard ML, OCaml, and Coq version 8's Gallina
sublanguage~\citep{SML,OCaml,Coq}.
I will assume that the reader is able to read basic Haskell code, and spend most
time explaining differences thereof.

\subsection{Lexical structure}

@@

On its own, an underscore has a completely different meaning, which can depend
on context.
In patterns, an underscore has the same meaning as it has in Haskell and ML ---
it holds the place of a pattern variable, but does not name that variable.
In expressions, an underscore stands for an unspecified subterm which will be
solved by unification~\citep{AP11,Miller92}.
The solving of unspecified terms is canonical and respects $\beta\eta$-equality,
unlike in Coq.

@@

An unfortunate feature of the definition \AgdaFunction{id$_0$} is that we cannot
apply it to the expression \AgdaPrimitive{Set}, because \AgdaPrimitive{Set}
contains only small types, and itself is a large type.
We can work around these size issues using
\emph{universe level polymorphism}~\citep{BCDE22},
as in the following definition.

@@

Slightly more flexible than parameters are \emph{Protestant indices}%
\footnote{The terminology of Protestant/Catholic indices is due to Peter
  Hancock. The mnemonic is that Catholics believe in transubstantiation, which
  is seen as analogous to the instantiation of Catholic indices with expressions
  that occurs during dependent pattern matching.}.
Protestant indices also appear to the left of the colon, and also must appear
unmodified in the \emph{return type} of all of the constructors.
However, they may take different values in inductive appearances of the type
family in the argument types of constructors.
Protestant indices give a generalisation of polymorphic
recursion to indices of arbitrary type~\citep{Mycroft84,Henglein93}.

@@

The second example of a type family with Catholic indices is more general in
nature.
Below I define \emph{propositional equality}, written
\AgdaDatatype{\_$\equiv$\_}.
It has two parameters and one Catholic index (though the standard library
version of propositional equality I use throughout this thesis has an extra
level parameter for the sake of universe level polymorphism).
The constructor \AgdaInductiveConstructor{refl} constructs an inhabitant of
\AgdaArgument{M}\AgdaSpace{}\AgdaDatatype{$\equiv$}\AgdaSpace{}%
\AgdaArgument{N} only when \AgdaArgument{N} is definitionally equal to
\AgdaArgument{M} (because terms are considered ``the same'' to the type checker
exactly when they are definitionally equal).
Notice that \AgdaInductiveConstructor{refl} does not quantify over \AgdaBound{x}
because \AgdaBound{x} is already in scope as a parameter.

### 2023-10-29 tex/dill.tex
[ writing ] ready for submission

The new feature when dealing with linear logic, compared to modal logic, is that
linear and intuitionistic assumptions satisfy different structural rules.
In our de Bruijn index style, where every multi-premise rule does the maximal
contraction and weakening occurs in the leafwardmost possible position,
this means that all of the rules have to manage the substructurality of linear
assumptions.
To help manage linear assumptions, I introduce three pieces of notation.

@@

\begin{figure}
  \begin{align*}
    A, B, C &\Coloneqq X \mid I \mid A \otimes B \mid A \multimap B
              \mid 0 \mid A \oplus B \mid \top \mid A \with B \mid \oc A \\
    \Gamma, \Delta, \Theta &\Coloneqq {\cdot} \mid \Gamma, A~\mathit{lin}
                             \mid \Gamma, A~\mathit{int} \\
    \mathcal S &\Coloneqq \Gamma \vdash A~\mathit{lin}
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \hypo{\Gamma \ni A}
      \infer1[Var]{\Gamma \vdash A~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash I~\mathit{lin}}
      \hypo{\Delta \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma~\mathit{int}}
      \infer1[$I$-I]{\Gamma \vdash I~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \otimes B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin}, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{int}}
      \hypo{\Delta \vdash B~\mathit{int}}
      \infer2[$\otimes$-I]{\Gamma + \Delta \vdash A \otimes B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \multimap B~\mathit{int}}
      \hypo{\Delta \vdash A~\mathit{int}}
      \infer2[$\multimap$-E]{\Gamma + \Delta \vdash B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma, A~\mathit{lin} \vdash B~\mathit{int}}
      \infer1[$\multimap$-I]{\Gamma \vdash A \multimap B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash 0~\mathit{lin}}
      \infer1[$0$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \text{(no \TirName{$0$-I})}
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \oplus B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin} \vdash C~\mathit{lin}}
      \hypo{\Delta, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer3[$\oplus$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A_i~\mathit{lin}}
      \infer1[$\oplus$-I$_i$]{\Gamma \vdash A_0 \oplus A_1~\mathit{lin}}
    }
    \and
    \text{(no \TirName{$\top$-E})}
    \and
    \ebrule{%
      \infer0[$\top$-I]{\Gamma \vdash \top~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A_0 \with A_1~\mathit{lin}}
      \infer1[$\with$-E$_i$]{\Gamma \vdash A_i~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma \vdash B~\mathit{lin}}
      \infer2[$\with$-I]{\Gamma \vdash A \with B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash \oc A~\mathit{lin}}
      \hypo{\Delta, A~\mathit{int} \vdash \oc C~\mathit{lin}}
      \infer2[$\oc$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma~\mathit{int}}
      \infer2[$\oc$-I]{\Gamma \vdash \oc A~\mathit{lin}}
    }
  \end{mathpar}
  \caption{Dual Intuitionistic Linear Logic}
  \label{fig:dill}
\end{figure}

>>>>

The new feature when dealing with linear logic, compared to modal logic, is that
linear and intuitionistic assumptions satisfy different structural rules.
In our De Bruijn index style, where every multi-premise rule does the maximal
contraction and weakening occurs in the leafwardmost possible position,
this means that all of the rules have to manage the substructurality of linear
assumptions.
To help manage linear assumptions, I introduce three pieces of notation.

@@

\begin{figure}
  \begin{align*}
    A, B, C &\Coloneqq X \mid I \mid A \otimes B \mid A \multimap B
              \mid 0 \mid A \oplus B \mid \top \mid A \with B \mid \oc A \\
    \Gamma, \Delta, \Theta &\Coloneqq {\cdot} \mid \Gamma, A~\mathit{lin}
                             \mid \Gamma, A~\mathit{int} \\
    \mathcal S &\Coloneqq \Gamma \vdash A~\mathit{lin}
  \end{align*}
  \begin{mathpar}
    \ebrule{%
      \hypo{\Gamma \ni A}
      \infer1[Var]{\Gamma \vdash A~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash I~\mathit{lin}}
      \hypo{\Delta \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma~\mathit{int}}
      \infer1[$I$-I]{\Gamma \vdash I~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \otimes B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin}, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer2[$I$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Delta \vdash B~\mathit{lin}}
      \infer2[$\otimes$-I]{\Gamma + \Delta \vdash A \otimes B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \multimap B~\mathit{lin}}
      \hypo{\Delta \vdash A~\mathit{lin}}
      \infer2[$\multimap$-E]{\Gamma + \Delta \vdash B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma, A~\mathit{lin} \vdash B~\mathit{lin}}
      \infer1[$\multimap$-I]{\Gamma \vdash A \multimap B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash 0~\mathit{lin}}
      \infer1[$0$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \text{(no \TirName{$0$-I})}
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A \oplus B~\mathit{lin}}
      \hypo{\Delta, A~\mathit{lin} \vdash C~\mathit{lin}}
      \hypo{\Delta, B~\mathit{lin} \vdash C~\mathit{lin}}
      \infer3[$\oplus$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A_i~\mathit{lin}}
      \infer1[$\oplus$-I$_i$]{\Gamma \vdash A_0 \oplus A_1~\mathit{lin}}
    }
    \and
    \text{(no \TirName{$\top$-E})}
    \and
    \ebrule{%
      \infer0[$\top$-I]{\Gamma \vdash \top~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A_0 \with A_1~\mathit{lin}}
      \infer1[$\with$-E$_i$]{\Gamma \vdash A_i~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma \vdash B~\mathit{lin}}
      \infer2[$\with$-I]{\Gamma \vdash A \with B~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash \oc A~\mathit{lin}}
      \hypo{\Delta, A~\mathit{int} \vdash \oc C~\mathit{lin}}
      \infer2[$\oc$-E]{\Gamma + \Delta \vdash C~\mathit{lin}}
    }
    \and
    \ebrule{%
      \hypo{\Gamma \vdash A~\mathit{lin}}
      \hypo{\Gamma~\mathit{int}}
      \infer2[$\oc$-I]{\Gamma \vdash \oc A~\mathit{lin}}
    }
  \end{mathpar}
  \caption{Dual Intuitionistic Linear Logic}
  \label{fig:dill}
\end{figure}

### 2023-10-29 tex/examples.tex
[ writing ] ready for submission

We can now use the elaborator to automatically infer the usage
annotations for the example at the end of \cref{sec:terms-of-system}.
This allows us to write:
\ExecuteMetaData[\PaperExamplestex]{cojoin}
We have instantiated the usage elaborator so that:
\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
this splitting would introduce more non-determinism but not allow any
more terms to be typed. As such, the only non-determinism comes when
we have variables annotated $\gr1$ and need to do an additive split,
like when we apply the \AgdaInductiveConstructor{!E} rule below. At
this point, the variable can become either $\gr0$-annotated in the
left subterm and $\gr1$-annotated on the right, or vice-versa. We will
find that, because the left subterm wants to use that variable, the
former choice will be rejected. The function \AgdaFunction{var\#} is a
convenience for converting statically known natural numbers,
representing de Bruijn \emph{levels}, into variable terms.

@@

@@

Now I give a \AgdaRecord{Semantics}.
The choice of \AgdaBound{$\V$} as
\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
given that a standard denotational semantics would not use intermediate
environments in the same sense as renaming and substitution, but it allows us to
reuse the standard facts that variables support renaming and identity
environments.
With this choice of \AgdaBound{$\V$} and \AgdaBound{$\C$}, we interpret
environment entries by \AgdaFunction{lookup$^R$}.

@@

>>>>

We can now use the elaborator to automatically infer the usage
annotations for the example at the end of \cref{sec:terms-of-system}.
This allows us to write:
\ExecuteMetaData[\PaperExamplestex]{cojoin}
We have instantiated the usage elaborator so that:
\AgdaField{0\#$^{-1}$} is a singleton on $\gr0$ and $\gr\omega$, and
empty on $\gr1$; \AgdaField{1\#$^{-1}$} is a singleton on $\gr1$ and
$\gr\omega$, and empty on $\gr0$; \AgdaField{+$^{-1}$} gives $\gr0
\mapsto [(\gr0,\gr0)]$, $\gr1 \mapsto [(\gr0,\gr1),(\gr1,\gr0)]$, and
$\gr\omega \mapsto [(\gr\omega,\gr\omega)]$; and \AgdaField{*$^{-1}$}
gives $(\gr\omega, \gr0) \mapsto [\gr0]$, $(\gr\omega, \gr1) \mapsto
[]$, and $(\gr\omega, \gr\omega) \mapsto [\gr\omega]$ (omitting
$(\gr0, \_)$ and $(\gr1, \_)$ cases for brevity). Note that we do not
consider splitting $\gr\omega$ up as, say, $\gr1 + \gr\omega$, because
this splitting would introduce more non-determinism but not allow any
more terms to be typed. As such, the only non-determinism comes when
we have variables annotated $\gr1$ and need to do an additive split,
like when we apply the \AgdaInductiveConstructor{!E} rule below. At
this point, the variable can become either $\gr0$-annotated in the
left subterm and $\gr1$-annotated on the right, or vice-versa. We will
find that, because the left subterm wants to use that variable, the
former choice will be rejected. The function \AgdaFunction{var\#} is a
convenience for converting statically known natural numbers,
representing De Bruijn \emph{levels}, into variable terms.

@@

Given the linear theme of $\name$, one may consider using the size-based
normalisation proof given by \citet[p.\ 71]{girard87linear} and
\citet{Krishnaswami13}.
However, this fails for two reasons for $\name$, both of which also apply to the
fragment I consider here defined by
$A, B \Coloneqq \iota \mid \oc\gr rA \multimap B$.
Firstly, we can use the $\oc$ connective in the linear instantiation of $\name$
to introduce contractable variables.
The size maintenance/reduction principles do not hold in the presence of
contraction, because substitution for a contracted variable can make the term
grow arbitrarily larger.
Secondly, there are instantiations other than the linear
$\{\gr0, \gr1, \gr\omega\}$ semiring, many of which allow for non-linear
behaviour even without the $\oc$-connective.
For example, with the trivial semiring, we recover an intuitionistic calculus
with no usage restrictions.
We therefore have to use a normalisation proof which also works for standard
Simply Typed $\lambda$-Calculus.

@@

Now I give a \AgdaRecord{Semantics}.
The choice of \AgdaBound{$\V$} as
\AgdaRecord{\AgdaUnderscore{}$\sqni$\AgdaUnderscore{}} is somewhat arbitrary,
given that a standard denotational semantics would not use intermediate
environments in the same sense as renaming, substitution, and evaluation.
The only requirements are that the choice of \AgdaBound{$\V$} respects renaming
and yields inhabitants of the target family
\AgdaFunction{$\llbracket$\_$\vdash$\_$\rrbracket$}.
Picking \AgdaRecord{\_$\sqni$\_} means that we can use the already proven lemma
\AgdaFunction{ren\textasciicircum$\sqni$} for the former requirement; and we use
\AgdaFunction{lookup$^R$} --- which we would have to provide somewhere anyway
--- to yield semantic objects.

@@

\section{Conclusion}

In this chapter, I have provided a range of example syntaxes and operations on
them, defined with the help of the generic semantic traversal.
Where \citet{AACMM21} give examples which focus on using the semantic
environment effectively, my examples focus more on problems we can only work on
using usage restrictions.
These examples tie together the two strands of this thesis --- representation of
syntax and semantics in Agda and usage restrictions via semiring annotations.

Though all of the applications have been successfully completed, there are some
points in them where details of the framework have forced us into an unnatural
or suboptimal approach, relative to what we could achieve with a
reimplementation of each specific calculus from first principles.
Specifically, in the NbE example of \cref{sec:nbe}, I was significantly hampered
by the inability to cleanly exclude variables from the syntax of normal forms,
and in the encoding of a linear/non-linear calculus in \cref{sec:lnl}, we were
forced to consider variables with linear type but intuitionistic usage
annotation and vice versa.
I will discuss these problems further, with suggestions of solutions, in
\cref{sec:conc}.

### 2023-10-29 tex/framework.tex
[ writing ] ready for submission

The work in this chapter is fully mechanised in Agda, which allows me to be
precise about the various levels of domain-specific languages which appear.

>>>>

The work in this chapter is fully mechanised in Agda, which allows me to be
precise about the various levels of domain-specific languages which appear.
Most of the code and text of this chapter, \cref{sec:semantics}, and the
examples in \cref{sec:usage-elaborator,sec:den-sem} is adapted from
\citet{WA22}.
The updated code for all of the remainder of this thesis is available at
\url{https://github.com/laMudri/generic-lr/releases/tag/thesis-submission}, and
can be cited via \doi{10.5281/zenodo.10050850}.

### 2023-10-29 tex/intro.tex
[ writing ] ready for submission

This thesis advances the frontier of programming languages work that can be done
in a proof assistant.
I will elaborate on both of these aspects in the following paragraphs.

The main aim of programming language research is to find common patterns in
computer programs and represent these patterns via programming language
features.
As a basic example, consider functions in, say, a low level language like C, in
contrast to what these functions are compiled to in assembly language/machine
code.
In order to support function calls, the C runtime system manages a
\emph{call stack}.
A function call is compiled to a sequence of instructions which store the return
address and the values of the arguments on the stack, move the stack pointer,
and jump to the compiled code of the function.
A return from a function then places the return value into the appropriate
register, moves the stack pointer back, and jumps back to the return address.
Advances in programming language techniques are evaluated by the usefulness of
behaviours captured and goodness of the mathematical properties of the
abstractions produced.

Within programming language research, type theory is a methodology for enforcing
abstractions by classifying program behaviour.
The simplest and most common type systems classify programs by what values they
may return.
For example, the body of a C function with return type \texttt{char} is a
program that, if it returns a value, will return a character.
If that function has a parameter of type \texttt{int}, then we can compose it
with a function with return type \texttt{int} to build up a larger program.
We expect type systems to stop us from running programs with arguments of the
wrong types.
This holds of both static and dynamic type systems --- with a static type
system, the compiler will refuse to compile our code if we pass a \texttt{char}
value to a function expecting an \texttt{int}, while with a dynamic type system,
our program will do a check before running the function to make sure that we
have passed it an \texttt{int}.
The abstraction produced in such simple type systems is the idea that a function
will take arguments in accordance with its parameter types and produce a result
in accordance with its result type, and furthermore, as readers, we do not have
to inspect the function's implementation to see that these properties are true.

A more recent trend in type theory is to use types to describe other parts of a
program's behaviour than what range of values it may return.
For example, Java's \emph{checked exceptions} can be seen as part of the static
type system, in which programs are classified by which exceptions they may
throw.
More generally, \emph{effect systems} classify programs based on all of the
effects they may have --- such as reading input and writing to files, and also
internally defined effects, such as non-determinism and using an accumulator.
In a type system which tracks effects (an \emph{effect system}), programs by
default are pure (i.e.\ have no effects), with effects being opt-in.
Pure programs generally enjoy good properties, making them easy to reason about
and easy for an optimising compiler to optimise.

@@

A \emph{proof assistant}, also known as an \emph{interactive theorem prover}, is
a piece of software that allows for the encoding of mathematical definitions,
theorems, constructions, and proofs, and furthermore check that such encoded
proofs are correct and that such encoded constructions are well formed.
To truly by interactive, i.e.\ to actually assist, a proof assistant will
usually have a user interface which can read partial proofs, display
information about what more proof needs to be given, and provide actions that
will help complete the proof.

@@

This thesis is written in \colour{}, but should be readable without.
Agda code has syntax highlighting, and various pieces of notation related to
usage annotations are coloured \gr{green} for emphasis.

>>>>

Programming language design has a history stretching back almost to the
inception of electronic computing.
Programming language \emph{theory} has arguably an even longer history, with
parts of mathematical logic developed in the early 20th century forming the
vocabulary of programming language theory as we know it today.
Programming language theory studies mathematical models of programming languages
to help us better understand programs in existing languages, and to inform the
design of new language features, new languages, and new programming paradigms.

%The field of programming language theory expands on at least two axes.
%On one axis, we become ever more able to give multiple mathematical
%interpretations of realistic programming language features.
%On another, the field generates new ideas with a theoretical basis which turn
%out to give rise to useful programming language features and paradigms.
%Most work pushes along both of these axes, including the work covered in this
%thesis, though most of the work covered in this thesis is phrased and directed
%in a way that emphasises the theoretical properties over the specifics of
%applications.
%In phrasing things this way, I hope to impart clear intuitions about the scope
%of this work, given its level of generality.

%This thesis advances the frontier of programming languages work that can be done
%in a proof assistant.
%I will elaborate on both of these aspects in the following paragraphs.

%The main aim of programming language research is to find common patterns in
%computer programs and represent these patterns via programming language
%features.
%As a basic example, consider functions in, say, a low level language like C, in
%contrast to what these functions are compiled to in assembly language/machine
%code.
%In order to support function calls, the C runtime system manages a
%\emph{call stack}.
%A function call is compiled to a sequence of instructions which store the return
%address and the values of the arguments on the stack, move the stack pointer,
%and jump to the compiled code of the function.
%A return from a function then places the return value into the appropriate
%register, moves the stack pointer back, and jumps back to the return address.
%Advances in programming language techniques are evaluated by the usefulness of
%behaviours captured and goodness of the mathematical properties of the
%abstractions produced.

Within programming language research, type theory is a methodology for
classifying program behaviour.
The simplest and most common type systems classify programs by what values they
may return.
For example, in many --- perhaps most --- widely used programming languages, the
compiler will keep track of the types of expressions and their subexpressions,
and give an error or warning to the programmer whenever there is a mismatch
between the expected type and the actual type.
For example, the body of a C function with return type \texttt{char} is a
program that, if it returns a value, will return a character.
If that function has a parameter of type \texttt{int}, then we can compose it
with a function with return type \texttt{int} to build up a larger program.
We expect type systems to stop us from running programs with arguments of the
wrong types.
This holds of both static and dynamic type systems --- with a static type
system, the compiler will refuse to compile our code if we pass a \texttt{char}
value to a function expecting an \texttt{int}, while with a dynamic type system,
our program will do a check before running the function to make sure that we
have passed it an \texttt{int}.
The abstraction produced in such simple type systems is the idea that a function
will take arguments in accordance with its parameter types and produce a result
in accordance with its result type, and furthermore, as readers, we do not have
to inspect the function's implementation to see that these properties are true.

A more recent trend in type theory is to use types to describe other parts of a
program's behaviour than what range of values it may return.
For example, Java's \emph{checked exceptions} can be seen as part of the static
type system, in which programs are classified by which exceptions they may
throw.
More generally, \emph{effect systems}~\citep{LG88} classify programs based on
all of the effects they may have ---
such as reading input and writing to files, and also
internally defined effects, such as non-determinism and using an accumulator.
In a type system which tracks effects (an \emph{effect system}), programs by
default are pure (i.e.\ have no effects), with effects being opt-in.
Pure programs generally enjoy good properties, making them easy to reason about
and easy for an optimising compiler to optimise.

@@

A \emph{proof assistant}, also known as an \emph{interactive theorem prover}, is
a piece of software that allows for the encoding of mathematical definitions,
theorems, constructions, and proofs, and furthermore check that such encoded
proofs are correct and that such encoded constructions are well formed.
To truly be interactive, i.e.\ to actually assist, a proof assistant will
usually have a user interface which can read partial proofs, display
information about what more proof needs to be given, and provide actions that
will help complete the proof.

@@

This thesis is written in \colour{}, but should be readable without.
Agda code has syntax highlighting, and various pieces of notation related to
usage annotations are coloured \gr{green} for emphasis.
As a test, the word \emph{colour} in the first sentence of this paragraph should
appear with the first two letters red, the next two green, and the final two
blue, with the shades of red, green, and blue used for Agda code.
Additionally, the word \emph{green} in the second sentence is written in the
shade of green used for usage annotations.

### 2023-10-29 tex/kits.tex
[ writing ] ready for submission

While substitution has this extra role in a lot of the syntaxes with binding we
care about, variable-binding also significantly complicates the substitution
operation.
Substitution acts on the free variables of a term, replacing them by terms, but
binders mean that some subterms have \emph{more} free variables than our
original term.
This causes different challenges for different representations of terms.
For example, with named variables and shadowing, na\"{i}vely defined
substitution could fall foul of variable capture.
In our approach, based on de Bruijn indices, the difficulty is that an index $i$
outside a binder of $n$ variables corresponds to an index $n + i$ inside the
binder.
Therefore, when substituting under a binder, we must first increment any free
variables contained in terms we are substituting in, which is a form of
\emph{renaming}.
Renaming replaces each free variable by another free variable, and is a special
case of substitution.
We must, however, define renaming before substitution, so as to avoid the
definition of substitution being circular.
Renaming avoids a similar circularity because when renaming goes under a binder,
we only have to increment each variable being renamed in, rather than each
variable \emph{in each term} being substituted in.

>>>>

While substitution has this extra role in a lot of the syntaxes with binding we
care about, variable-binding also significantly complicates the substitution
operation.
Substitution acts on the free variables of a term, replacing them by terms, but
binders mean that some subterms have \emph{more} free variables than our
original term.
This causes different challenges for different representations of terms.
For example, with named variables and shadowing, na\"{i}vely defined
substitution could fall foul of variable capture.
In our approach, based on De Bruijn indices, the difficulty is that an index $i$
outside a binder of $n$ variables corresponds to an index $n + i$ inside the
binder.
Therefore, when substituting under a binder, we must first increment any free
variables contained in terms we are substituting in, which is a form of
\emph{renaming}.
Renaming replaces each free variable by another free variable, and is a special
case of substitution.
We must, however, define renaming before substitution, so as to avoid the
definition of substitution being circular.
Renaming avoids a similar circularity because when renaming goes under a binder,
we only have to increment each variable being renamed in, rather than each
variable \emph{in each term} being substituted in.

### 2023-10-29 tex/lenv.tex
[ writing ] full draft
[ writing ] ready for submission

We could, as in \cref{def:lr-rec-env}, use these three clauses to define what an
environment is.
However, I find them difficult to work with, as it is often easier to do
linear algebraic proofs separately from the rest of an environment.
For example, for identity and composition of environments, as we are about to
see, our working definition (\cref{def:lr-env})
is easier to use because we can rely on the closure of linear maps under
identity and composition.
Concretely, an inductive construction of, for example, an identity environment
would involve constructing an environment of type
$\grP\gamma, \grQ\delta \env\V \grP\gamma, \grQ\delta$ by constructing
environments of types $\grP\gamma, \gr0\delta \env\V \grP\gamma$ and
$\gr0\gamma, \grQ\delta \env\V \grQ\delta$.
These are not identity environments, so we would have to strengthen the
induction hypothesis.
Similar strengthening of the induction hypothesis would be required for
composition.

@@

I put together the preceeding pieces to give a syntactic traversal operation over
$\name$ in the following section.
For the rest of this section, I observe some more constructions purely on
environments --- in particular, composition of environments given certain
assumptions about the families of values.

>>>>

We could, as in \cref{def:lr-rec-env}, use these three clauses to define what an
environment is.
However, such a definition appears to require creative induction hypotheses in
the proving of simple lemmas, in contrast to the more direct proofs I achieve
below using \cref{def:lr-env}.
To take a concrete example, consider how we may construct an ``identity''
environment of type $\Gamma \env\V \Gamma$, as in \cref{thm:env-id} below.
If we try to directly proceed by induction on $\Gamma$, we get to the case where
we are aiming to construct an environment of type
$\grP\gamma, \grQ\delta \env\V \grP\gamma, \grQ\delta$ by constructing
environments of types $\grP\gamma, \gr0\delta \env\V \grP\gamma$ and
$\gr0\gamma, \grQ\delta \env\V \grQ\delta$.
These are not identity environments, and thus do not come from the hypotheses of
a simple induction.
In contrast, using \cref{def:lr-env}, in \cref{thm:env-id} we are able to use
the standard fact that there are identity linear maps, and on top of such a map
worry only about the value assigned to each variable.

@@

I put together the preceding pieces to give a syntactic traversal operation over
$\name$ in the following section.
For the rest of this section, I observe some more constructions purely on
environments --- in particular, composition of environments given certain
assumptions about the families of values.
